Adversarially Robust Medical Classification via Attentive Convolutional Neural Networks

### 2022-07-27 paper/PaperForReview.tex
Reorganized figures

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{lipsum}
\usepackage{tikz}
\usepackage{relsize}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{subcaption}
% \usepackage{float}
\pgfplotsset{compat=1.7}

@@

      \begin{figure}[h!]
        \centering
        $
        \begin{array}{l}
        \includegraphics[width=0.4\linewidth]{graphics/ResNet-50/unprocessedxray.png}
        \end{array}
        \mathlarger{\mathlarger{\mathlarger{\mathlarger{\rightarrow}}}}
        \begin{array}{l}
          \includegraphics[width=0.4\linewidth]{graphics/ResNet-50/processedxray.jpg}
        \end{array}
        $
        \caption{Visualization of the chest x-ray image preprocessing pipeline}
      \end{figure}

@@

@@

  \section{Results}
    \subsection{ResNet-50 Models}
      For the task of skin lesion classification, the models with attention are clearly superior, in terms of robustness (Fig. \ref{DermResNet50Robustness}). Although the model without attention has slightly higher clean-accuracy (0.905 vs. 0.88), even the slightest perturbation ($\epsilon=0.00125$) is able to reduce its accuracy by 24\% and put the model with attention in the lead. By the time $\epsilon=0.005$, the model is worse than random, and by the time $\epsilon=0.01$, the model is rendered completely useless. It is worth noting that perturbation of this size (Fig. \ref{example_derm_perturbation}) is far from being human perceptible. Meanwhile, at this perturbation radius, the model with attention remains better than random selection.

      \begin{figure}[h!]
        \begin{tikzpicture}
          \begin{axis}[xtick={0, 0.00125, 0.0025, 0.005, 0.01, 0.02, 0.04, 0.08, 0.16, 0.32}, x tick label style={rotate=45, log ticks with fixed point},xmode=log, log basis x=2, xlabel=Perturbation Radius ($\epsilon$), ylabel=Accuracy, width=\linewidth, height=7cm,legend style={at={(0.95,0.95)},anchor=north east}]

          \addplot[color=red,mark=x] coordinates {
            (0.00125, 0.687)
            (0.0025, 0.216)
            (0.005, 0.012)
            (0.01, 0)
          };

          \addplot[color=blue,mark=*] coordinates {
            (0.00125, 0.795)
            (0.0025, 0.685)
            (0.005, 0.337)
            (0.01, 0.022)
          };

          \addplot[color=red, domain=0.00125:0.01, dashed]{0.905};
          \addplot[color=blue, domain=0.00125:0.01, dashed]{0.880};

          \legend{Without Attention,With Attention}
          \end{axis}
          \end{tikzpicture}
        \caption{Clean- and robust-accuracy of ResNet-50 models for skin lesion classification.  Dashed lines represent clean-accuracy.}
        \label{DermResNet50Robustness}
      \end{figure}

@@

      \begin{figure}[h!]
        \begin{tikzpicture}
          \begin{axis}[xtick={0, 0.00125, 0.0025, 0.005, 0.01, 0.02, 0.04, 0.08, 0.16, 0.32}, x tick label style={rotate=45, log ticks with fixed point},xmode=log, log basis x=2, xlabel=Perturbation Radius ($\epsilon$), ylabel=Accuracy, width=\linewidth, height=7cm,legend style={at={(0.05,0.05)},anchor=south west}]

          \addplot[color=red,mark=x] coordinates {
            (0, 0.832)
            (0.00125, 0.828)
            (0.0025, 0.825)
            (0.005, 0.819)
            (0.01, 0.806)
            (0.02, 0.778)
            (0.04, 0.705)
            (0.08, 0.506)
            (0.16, 0.175)
            (0.32, 0.014)
          };

          \addplot[color=blue,mark=*] coordinates {
            (0, 0.818)
            (0.00125, 0.816)
            (0.0025, 0.815)
            (0.005, 0.813)
            (0.01, 0.808)
            (0.02, 0.8)
            (0.04, 0.780)
            (0.08, 0.733)
            (0.16, 0.599)
            (0.32, 0.287)
          };

          \addplot[color=red, domain=0.00125:0.32, dashed]{0.832};
          \addplot[color=blue, domain=0.00125:0.32, dashed]{0.818};

          \legend{Without Attention,With Attention}
          \end{axis}
          \end{tikzpicture}
        \caption{Clean- and robust-accuracy of ResNet-50 models for diabetic retinopathy detection. Dashed lines represent clean-accuracy.}
        \label{DRResNet50Robustness}
      \end{figure}

@@

      \begin{figure}[h!]
        \begin{subfigure}{\linewidth}
          \centering
          $
          \begin{array}{l}
          \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/derm_unperturbed.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{+}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/derm_e=0.01_difference.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{=}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/derm_e=0.01.jpg}
          \end{array}
          $
          \caption{An example of $\epsilon=0.01$ perturbation on the dermatoscopy dataset. 0.01 is the maximum perturbation radius used for this dataset.\\}
          \label{example_derm_perturbation}
        \end{subfigure}

        \begin{subfigure}{\linewidth}
          \centering
          $
          \begin{array}{l}
          \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/dr_unperturbed.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{+}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/dr_e=0.32_difference.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{=}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/dr_e=0.32.jpg}
          \end{array}
          $
          \caption{An example of $\epsilon=0.32$ perturbation on the fundoscopy dataset. 0.32 is the maximum perturbation radius used for this dataset.\\}
          \label{example_dr_perturbation}
        \end{subfigure}
        \begin{subfigure}{\linewidth}
          \centering
          $
          \begin{array}{l}
          \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/pneumo_unperturbed.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{+}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/pneumo_e=0.32_difference.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{=}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/ResNet-50/pneumo_e=0.32.jpg}
          \end{array}
          $
          \caption{An example of $\epsilon=0.32$ perturbation on the chest x-ray dataset. 0.32 is the maximum perturbation radius used for this dataset.\\}
          \label{example_pneumo_perturbation}
        \end{subfigure}%
        \caption{Perturbation examples}
      \end{figure}

      \begin{figure}[h!]
        \begin{tikzpicture}
          \begin{axis}[xtick={0, 0.00125, 0.0025, 0.005, 0.01, 0.02, 0.04, 0.08, 0.16, 0.32}, x tick label style={rotate=45, log ticks with fixed point},xmode=log, log basis x=2, xlabel=Perturbation Radius ($\epsilon$), ylabel=Accuracy, width=\linewidth, height=7cm,legend style={at={(0.05,0.05)},anchor=south west}]

          \addplot[color=red,mark=x] coordinates {
            (0.00125, 0.897)
            (0.0025, 0.897)
            (0.005, 0.896)
            (0.01, 0.896)
            (0.02, 0.894)
            (0.04, 0.889)
            (0.08, 0.881)
            (0.16, 0.864)
            (0.32, 0.805)
          };

          \addplot[color=blue,mark=*] coordinates {
            (0.00125, 0.883)
            (0.0025, 0.883)
            (0.005, 0.882)
            (0.01, 0.882)
            (0.02, 0.880)
            (0.04, 0.877)
            (0.08, 0.870)
            (0.16, 0.853)
            (0.32, 0.805)
          };

          \addplot[color=red, domain=0.00125:0.32, dashed]{0.897};
          \addplot[color=blue, domain=0.00125:0.32, dashed]{0.883};

          \legend{Without Attention,With Attention}
          \end{axis}
          \end{tikzpicture}
        \caption{Clean- and robust-accuracy of ResNet-50 models for pneumothorax detection. Dashed lines represent clean-accuracy.}
        \label{PneumoResNet50Robustness}
      \end{figure}

    \subsection{InceptionResNetV2 Models}
  \section{Discussion}
    \subsection{Analysis of Perturbations and Activation Maps}
      In an attempt to better understand the principles and behaviors that led the CNNs with attention to perform better in both clean and adversarial scenarios, perturbation difference maps and Grad-CAM activation maps were generated for a select sample of images on each model. Individually, these difference maps and activation maps were unremarkable. However, a number of dataset specific patterns were noticed throughout the images used for this analysis.
      \subsubsection{ResNet-50 Models}
        While imperceptible in the final images, the perturbations for dermatoscopic and fundoscopic images were found to carry easily perceptible information about the source image, when the changes were scaled to a range of $[0,255]$. In these cases, the adversarial ``noise'' contained the shape of the lesion or eye. For dermatoscopic images, this shape was in the form of a densely perturbed ring enclosing a sparsely perturbed center (Fig. \ref{LesionShape}). This phenomenon was most visible when $\epsilon=0.00125$. For the fundoscopic images, this shape was represented by a sparsely perturbed ellipse. In the attacks generated for the model with attention, the shape of this ellipse occasionally diverged from the true shape of the eye (Fig. \ref{EyeShapeDivergence}).

        \begin{figure}[h]
          \centering
          \begin{subfigure}{.4\linewidth}
            \centering
            \includegraphics[width=\linewidth]{graphics/ResNet-50/LesionOutline.pdf}
            \caption{Shape of skin lesion found in adversarial ``noise''\\}
            \label{LesionShape}
          \end{subfigure}
          \begin{subfigure}{.4\linewidth}
            \centering
            \includegraphics[width=\linewidth]{graphics/ResNet-50/EyeShapeDivergence.pdf}
            \caption{Divergence from shape of eye\\}
            \label{EyeShapeDivergence}
          \end{subfigure}
          \caption{Shape information carried over into perturbations}
        \end{figure}

        \begin{figure}[h]
          \centering
          \includegraphics[width=\linewidth]{graphics/ResNet-50/pneumo_with_attention.pdf}
          \caption{Comparison of activation maps (with attention) for perturbed and unperturbed chest x-ray}
          \label{PneumoWithAttention}
        \end{figure}

        \begin{figure}[h]
          \centering
          \includegraphics[width=\linewidth]{graphics/ResNet-50/different_maps.pdf}
          \caption{Comparison of activation maps (without attention) for perturbed and unperturbed chest x-ray}
          \label{DifferentMaps}
        \end{figure}

@@

        \begin{figure}[h]
          \centering
          \includegraphics[width=0.5\linewidth]{graphics/ResNet-50/cotton_wool.pdf}
          \caption{Dark splotches in perturbation of fundoscopic images}
          \label{CottonWool}
        \end{figure}

>>>>

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{lipsum}
\usepackage{tikz}
\usepackage{relsize}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{subcaption}
\usepackage{multicol}
% \usepackage{float}
\pgfplotsset{compat=1.7}

@@

      \input figures/XRayProcessingPipeline

@@

      \input figures/ExamplePerturbations

@@

  \section{Results}
    \subsection{ResNet-50 Models}
      \input figures/ResNetDermPlot
      \input figures/ResNetDRPlot
      \input figures/ResNetPneumoPlot

      For the task of skin lesion classification, the models with attention are clearly superior, in terms of robustness (Fig. \ref{DermResNet50Robustness}). Although the model without attention has slightly higher clean-accuracy (0.905 vs. 0.88), even the slightest perturbation ($\epsilon=0.00125$) is able to reduce its accuracy by 24\% and put the model with attention in the lead. By the time $\epsilon=0.005$, the model is worse than random, and by the time $\epsilon=0.01$, the model is rendered completely useless. It is worth noting that perturbation of this size (Fig. \ref{example_derm_perturbation}) is far from being human perceptible. Meanwhile, at this perturbation radius, the model with attention remains better than random selection.

@@

@@

    \subsection{InceptionResNetV2 Models}

      \input figures/IRV2DermPlot
      \input figures/IRV2DRPlot
      \input figures/IRV2PneumoPlot

      Experiments with the InceptionResNetV2 architecture paint a slightly different picture. In all tasks, clean-accuracy for the baseline and attentive models was within 1\%. This result is, somewhat, at odds with those of Datta et al. (2021) \cite{AttentionSkinCancerClassification} which found soft attention to improve accuracy of InceptionResNetV2 (on HAM10000 \cite{HAM10000}) by ~3\%.

      For the skin lesion classification and diabetic retinopathy detection tasks, the baseline and attentive architectures performed near-identically in adversarial scenarios. The diabetic retinopathy task slightly favored the baseline, whereas skin lesion classification slightly favored the attentive model; however, at most, there was a 6\% discrepancy between their accuracies ($\epsilon = 0.00125$).

      In the pneumothorax detection task, the model with attention performed significantly worse under adversarial scenarios. The model reached an unusable accuracy after the perturbation radius was pushed beyond 0.00125, while the baseline remained usable with a perturbation radius of 0.005.

  \section{Discussion}
    \subsection{Analysis of Perturbations and Activation Maps}
      In an attempt to better understand the principles and behaviors that led the CNNs with attention to perform better in both clean and adversarial scenarios, perturbation difference maps and Grad-CAM activation maps were generated for a select sample of images on each model. Individually, these difference maps and activation maps were unremarkable. However, a number of dataset specific patterns were noticed throughout the images used for this analysis.
      \subsubsection{ResNet-50 Models}
        While imperceptible in the final images, the perturbations for dermatoscopic and fundoscopic images were found to carry easily perceptible information about the source image, when the changes were scaled to a range of $[0,255]$. In these cases, the adversarial ``noise'' contained the shape of the lesion or eye. For dermatoscopic images, this shape was in the form of a densely perturbed ring enclosing a sparsely perturbed center (Fig. \ref{LesionShape}). This phenomenon was most visible when $\epsilon=0.00125$. For the fundoscopic images, this shape was represented by a sparsely perturbed ellipse. In the attacks generated for the model with attention, the shape of this ellipse occasionally diverged from the true shape of the eye (Fig. \ref{EyeShapeDivergence}).

        \input figures/DRPerturbationDivergence
        \input figures/ResNetPneumoWithAttentionActivation
        \input figures/ResNetPneumoWithoutAttentionActivation

@@

        \input figures/ResNetCottonWool
